DeconvOptim.jl - Signal Deconvolution with Julia

### 2021-08-09 paper/paper.tex

\section{Introduction}
Deconvolution has been a long addressed problem especially in astronomical imaging, digital signal processing or microscopy imaging.
The assumption is that images are degraded by some kind of blur which can be described as a convolution of a kernel $h$ with the image $S$:
\begin{equation}
    I(\mathbf r) = (h * S)(\mathbf r)
\end{equation}
where $*$ denotes a convolution operation.
$h$ is often called the point spread function (PSF) since it characterizes how the systems maps a point in the object to a blurred spot 
in the image.
Often this convolution operation is disturbed by different types of noise where the most common are additive Gaussian noise (e.g. due to the read out of the camera sensors) or Poisson shot noise (e.g. due to the quantum nature of photons):
\begin{align}
    Y_{\text{Poisson}}(\mathbf r) &= \text{Poisson}[(h * S)(\mathbf r)]\\
    Y_{\text{Gauss}}(\mathbf r) &= (h * S)(\mathbf r) + \text{Gauss}_{\sigma}(\mathbf r)
\end{align}
Due to noise and the low pass filter effect of the PSF, deconvolution is an ill-posed problem which cannot be solved directly.\\
A well known approach, because of its simplicity, has been the iterative Richardson-Lucy deconvolution algorithm 
\cite{Richardson:72}, \cite{lucy:74} for Poisson noise degraded signals.
Another common method is the non-iterative Wiener filter \cite{wiener2013extrapolation} because of its predictable runtime.
The Wiener filter assumes additive noise which is often not a good approximation, especially in low light conditions as in microscopy or astronomy. 
Recently, machine learning based methods have become more and more favoured in deconvolution because
of the possibility to add learned properties of similar images to the algorithm \cite{Kruse_2017_ICCV}.
However, those implementations are usually restricted to cases where experimental datasets are available for the characterization and training.
\verb|DeconvOptim.jl| is a toolbox written in Julia \cite{bezanson2017julia} which offers several deconvolution algorithms 
like the iterative Richardson-Lucy algorithm but also
much more flexible methods which are based on minimizing a 
noise-dependent loss function with the help of modern gradient-descent minimization routines. 
Also this toolbox is not restricted in dimensionality and therefore can be applied to deconvolution problems in many different fields.

@@

\subsection{FFT Based Convolution}
    The basis for fast convolution calculation via Fourier transforms lies in the convolution theorem. It states that
    \begin{equation}
        \mathcal F\left\{(h * S)\right\}(\mathbf k) = \mathcal F\{h\} (\mathbf k) \cdot \mathcal F\{S\} (\mathbf k),
        \label{eq:fftconv}
    \end{equation} 
        which means that a Fourier transform $\mathcal F$ of a convolution operation is equivalent to a point wise multiplication of the Fourier transforms 
        of $h$ and $S$.
        Fortunately, the Fourier transform operations can be calculated even for moderate large 3D arrays quickly on modern computing machines via
        the Fast Fourier Transform (FFT).
        In Julia, \verb|FFTW.jl| offers a convenient interface to the FFTW library \cite{FFTW05}.
        The time complexity to calculate \autoref{eq:fftconv} is $\mathcal{O}(N \cdot \log N)$ for a one dimensional discretized dataset with $N$ samples.
        For a 2D dataset with size $N \times M$ it is $\mathcal{O}(N \cdot M \log(N \cdot M))$ and accordingly for higher dimensions.
        Spatially based convolution kernels with size $K \times L$ applied on an image with
        $N \times M$ data points have a runtime of $\mathcal O(N \cdot M \cdot K \cdot L)$.
        It is clear that for larger kernel sizes ($K,L \gg 1$) the sliding method 
        has worse computational complexity than the FFT based convolutions. The many cores of a GPU can mitigate that to a certain extent
        but since the FFT (CuFFT in the CUDA library \cite{CUDA}) can be also executed in parallel fashion, 
        both approaches profit from parallelization.

@@

        However, without the ground truth image it is not clear which iteration or $\lambda$ is optimal and hence
        this depends on the user.

@@

    \subsection{Performance Comparison}
        Since Julia is often advertised with great performance we compare parts and the full deconvolution to other implementations in
        terms of their total performance.
        \subsubsection{Regularizer}
            In \autoref{tab:reg} we can see the runtimes for different TV regularizer implementations. As we can see, \verb|Tullio.jl| outperforms a plain 
            Julia implementation by more than order of magnitude. Partially, that is caused by \verb|Tullio.jl|
            being able to 
            use of all threads whereas Julia's built in broadcast mechanism is not multi threaded.
            By \textit{plain Julia} we are referring to an implementation which is written with standard Julia Base functions.
            The same plain Julia implementation was then used together with \verb|CUDA.jl|
            to make use of the GPU capabilities. We see that there is a minor benefit in using a GPU for the regularizer instead of 
            the \verb|Tullio.jl| version on the CPU. Unfortunately, GPU support by \verb|Tullio.jl| is still experimental and especially on the gradient step 
            we could not achieve reasonable results with it.
            \begin{table}[h]
                \begin{tabular}{l l l l l}
                    & plain & $\nabla$ plain & \verb|Tullio.jl| & $\nabla$ \verb|Tullio.jl| \\
                \hline
                CPU & \SI{83.9}{\milli\second} & \SI{7180}{\milli\second} & \SI{5.53}{\milli\second}  & \SI{119.8}{\milli\second}   \\ 
                GPU & \SI{2.5}{\milli\second}  & \SI{22.9}{\milli\second}& \SI{4.00}{\milli\second}  & \SI{12530}{\milli\second} \\
                \hline
                \hline
                    & JAX & $\nabla$ JAX && \\
                \hline
                CPU & \SI{48}{\milli\second}   & \SI{364}{\milli\second}&   & \\
                GPU & \SI{0.887}{\milli\second}   & \SI{10.2}{\milli\second}&   &
                \end{tabular}
                \caption{Runtimes of a single Total Variation regularizer calculation for different Julia versions and JAX. 
                    We used a 3 dimensional Float32 array with size $300 \times 300 \times 300$.  }
                \label{tab:reg}
            \end{table}
            We also compare the performance to an implementation written in Python with the Framework JAX \cite{jax2018github}. 
            The GPU version seems to outperform the Julia version by roughly a factor of $2 - 3$ in both forward and gradient pass.
            However, on the CPU \verb|Tullio.jl| is considerably faster. 
            In general one could always squeeze more performance out of the gradient calculations if the kernels would be hand optimized.
            In our package we especially try avoid that because of the multi-dimensionality it is much more convenient
            to rely on AD and as shown above the performance is usually acceptable.

    \subsection{Full Deconvolution}
        In \autoref{tab:deconv} we show the deconvolution time for a 3D dataset with a size of $512 \times 256 \times 128$\footnote{Dataset was taken from \url{http://bigwww.epfl.ch/deconvolution/data/microtubules/}\cite{deconvlab2}. During our experiments we noticed a pixel offset between the ground truth and the noisy image. To reproduce our results, one has to correct this offset if the images are compared quantitatively.} 
        The basic settings of the different algorithms and packages are noted in \autoref{tab:deconvprop}.
        The aim of this comparison is not to compare the quality of the results but the computing time to achieve similar reconstructions. 
        The algorithms used in \verb|DeconvOptim.jl| are not fundamentally different to a plain Richardson-Lucy deconvolution and therefore the results will 
        not exceed those results significantly in terms of quality.\\
        One should keep in mind that such performance comparisons are usually biased since we are not
        familiar with all options which should be chosen to achieve best performance.
        The comparison should give a rough overview which runtime is expected for each toolbox.\\
        The different packages used were \verb|ThreeDeconv.jl| \cite{ikoma2018convex}, \verb|Huygens| Professional version 21.04 (Scientific Volume Imaging, The Netherlands, \url{http://svi.nl}), \verb|DeconvolutionLab2| \cite{deconvlab2} and \verb|GenericDeconv|.

        \begin{table}[h]
            \begin{tabular}{l l l l }
                &  CPU in \si{\second}& GPU in \si{\second} & NCC\\ 
            \hline
            \verb|DeconvOptim.jl| & 32.9 & 2.20 & 0.86\\
            \verb|DeconvOptim.jl| - TV & 39.9 & 4.46 & 0.79\\
            Richardson-Lucy (Julia) & 40.0 & 3.05 & 0.74 \\
            \verb|ThreeDeconv.jl|& 1550 & 27.1 & 0.85\\
            \verb|GenericDeconv|  & 140 & 6.32 & 0.86\\
            \verb|GenericDeconv| - TV & 233 & 9.35 & 0.73\\
            \verb|Huygens| & 30.1 & 6.8 & 0.86\\
            \verb|DeconvolutionLab2| - LR& 1640 & & 0.76\\
            \end{tabular}
            \caption{Runtimes of different deconvolution implementations. See \autoref{tab:deconvprop} for the number of iterations
            for each method. We have chosen the same number of iterations for the deconvolutions with and without regularizers.}
            \label{tab:deconv}
        \end{table}

        For the results of \verb|GenericDeconv| (written in Matlab by Rainer Heintzmann) we used a Poisson loss function with a L-BFGS minimizer 
        and a quadratic non-negativity constraint.
        When the TV regularizer was added the weight was $\lambda = 0.001$.
        The general idea of \verb|GenericDeconv| is very similar to \verb|DeconvOptim.jl|.
        \verb|DeconvolutionLab2| offers a wide variety of different algorithms which however only run on a CPU. 
        In the table we show the results of an 
        iterative Richardson-Lucy deconvolution. The regularizer weight was $\lambda=0.001$.\\
        In \verb|DeconvOptim.jl| we measured the runtime of the default deconvolution routine with a Poisson loss function
        and the runtime of straightforward Richardson-Lucy deconvolution without a regularizer.\\
        \verb|ThreeDeconv.jl| is another Julia deconvolution toolbox for deconvolution of three dimensional data, the regularizer weight was $\lambda=10^{-7}$. 
        Since it is hard to judge the quality of those results with a numerical loss function value, 
        we compare them by the NCC, see \autoref{img:results}.
        The CPU versions were multi threaded (if possible). 
        \verb|DeconvolutionLab2|, \verb|Huygens| and \verb|GenericDeconv| were tested under Windows 10 and not Ubuntu.
        \todo{anscombe norm speed}
        \todo{Fix table }
        \begin{table}[h]
            \begin{tabular}{l l l}
                & Loss & iterations \\ 
            \hline
            \verb|DeconvOptim.jl| & Poisson & 45 \\
            \verb|DeconvOptim.jl| TV & Poisson & 45\\
            Richardson-Lucy & Poisson & 300 \\
            \verb|ThreeDeconv.jl|  & Gauss+Poisson& 300\\
            \verb|GenericDeconv|  & Poisson & 45\\
            \verb|GenericDeconv| TV & Poisson & 45 \\
            \verb|Huygens| & & \\
            \verb|DeconvolutionLab2| LR & Poisson & 300\\
            \end{tabular}
            \caption{Overview which loss function (noise model) and number of iterations are used.}
            \label{tab:deconvprop}
        \end{table}
        As summary of those performance tests we can state that our Julia implementation is very competitive and especially on the GPU
        it was the fastest of the tested packages. 
        We also see that our routine was faster than a plain Richardson-Lucy mainly due to the much lower number
        of iterations.
        On the CPU \verb|DeconvOptim.jl| is only beaten by the commercial \verb|Huygens| software. 
        Despite \verb|ThreeDeconv.jl| written in Julia it is notably slower in comparison to our implementation.
        \verb|DeconvolutionLab2| is written in Java and also uses hand written gradients but still its speed is poor in
        comparison to all other packages.

>>>>

\section{Introduction}
Deconvolution has been a long addressed problem especially in astronomical imaging, digital signal processing or microscopy imaging.
The assumption is that images are degraded by some kind of blur which can be described as a convolution of a kernel $h$ with the image $S$:
\begin{equation}
    I(\mathbf r) = (h * S)(\mathbf r)
\end{equation}
where $*$ denotes a convolution operation.
$h$ is often called the point spread function (PSF) since it characterizes how the systems maps a point in the object to a blurred spot 
in the image.
Often this convolution operation is disturbed by different types of noise where the most common are additive Gaussian noise (e.g. due to the read out of the camera sensors) or Poisson shot noise (e.g. due to the quantum nature of photons):
\begin{align}
    Y_{\text{Poisson}}(\mathbf r) &= \text{Poisson}[(h * S)(\mathbf r)]\\
    Y_{\text{Gauss}}(\mathbf r) &= (h * S)(\mathbf r) + \text{Gauss}_{\sigma}(\mathbf r)
\end{align}
Due to noise and the low pass filter effect of the PSF, deconvolution is an ill-posed problem which cannot be solved directly.\\
A well known approach, because of its simplicity, has been the iterative Richardson-Lucy deconvolution algorithm 
\cite{Richardson:72}, \cite{lucy:74} for Poisson noise degraded signals.
Another common method is the non-iterative Wiener filter \cite{wiener2013extrapolation} because of its predictable runtime.
The Wiener filter assumes additive noise which is often not a good approximation, especially in low light conditions as in microscopy or astronomy. 
Recently, machine learning based methods have become more and more favoured in deconvolution because
of the possibility to add learned properties of similar images to the algorithm \cite{Kruse_2017_ICCV}.
However, those implementations are usually restricted to cases where experimental datasets are available for the characterization and training.
\verb|DeconvOptim.jl| is a toolbox written in Julia \cite{bezanson2017julia} which offers several deconvolution algorithms 
like the iterative Richardson-Lucy algorithms but also
much more flexible methods which are based on minimizing a 
noise-dependent loss function with the help of modern gradient-descent minimization routines. 
Also this toolbox is not restricted in dimensionality and therefore can be applied to deconvolution problems in many different fields.

@@

\subsection{FFT Based Convolution}
    The basis for fast convolution calculation via Fourier transforms lies in the convolution theorem. It states that
    \begin{equation}
        \mathcal F\left\{(h * S)\right\}(\mathbf k) = \mathcal F\{h\} (\mathbf k) \cdot \mathcal F\{S\} (\mathbf k),
        \label{eq:fftconv}
    \end{equation} 
        which means that a Fourier transform of a  convolution operation is equivalent to a point wise multiplication of the Fourier transforms 
        of $h$ and $S$.
        Fortunately, the Fourier transform operations can be calculated even for moderate large 3D arrays quickly on modern computing machines via
        the Fast Fourier Transform (FFT).
        In Julia, \verb|FFTW.jl| offers a convenient interface to the FFTW library \cite{FFTW05}.
        The time complexity to calculate \autoref{eq:fftconv} is $\mathcal{O}(N \cdot \log N)$ for a one dimensional discretized dataset with $N$ samples.
        For a 2D dataset with size $N \times M$ it is $\mathcal{O}(N \cdot M \log(N \cdot M))$ and accordingly for higher dimensions.

\subsection{Sliding Convolution Windows}
Convolutions can be also calculated with a sliding kernel in real space which is the direct approach to
solve the integral in \autoref{eq:conv}.
That mechanism is the working principle
behind the convolutions in neural networks. The reason is that this operation can be parallelized easily and therefore 
is suited for GPUs. However, the computational complexity for a 2D kernel with size $K \times L$ applied on an image with
$N \times M$ data points is $\mathcal O(N \cdot M \cdot K \cdot L)$. It is clear that for larger kernel sizes ($K,L \gg 1$) the sliding method 
has worse computational complexity than the FFT based convolutions. The many cores of a GPU can mitigate that to a certain extent
but since the FFT (CuFFT in the CUDA library \cite{CUDA}) can be also executed in parallel fashion, both approaches profit from parallelization.
If the kernel is separable over the dimensions (like a Gaussian kernel) the complexity changes to $\mathcal O(N \cdot M \cdot L +  N \cdot M \cdot K)$ however most optical PSFs are not separable and therefore this is not generally applicable.

@@

@@

    \subsection{Performance Comparison}
        Since Julia is often advertised with great performance we compare parts and the full deconvolution to other implementations in
        terms of their total performance.
        \subsubsection{Regularizer}
            In \autoref{tab:reg} we can see the runtimes for different TV regularizer implementations. As we can see, \verb|Tullio.jl| outperforms a plain 
            Julia implementation by more than order of magnitude. Partially, that is caused by \verb|Tullio.jl|
            being able to 
            use of all threads whereas Julia's built in broadcast mechanism is not multi threaded.
            By \textit{plain Julia} we are referring to an implementation which is written with standard Julia Base functions.
            The same plain Julia implementation was then used together with \verb|CUDA.jl|
            to make use of the GPU capabilities. We see that there is a minor benefit in using a GPU for the regularizer instead of 
            the \verb|Tullio.jl| version on the CPU. Unfortunately, GPU support by \verb|Tullio.jl| is still experimental and especially on the gradient step 
            we could not achieve reasonable results with it.
            \begin{table}[h]
                \begin{tabular}{l l l l l}
                    & plain & $\nabla$ plain & \verb|Tullio.jl| & $\nabla$ \verb|Tullio.jl| \\
                \hline
                CPU & \SI{83.9}{\milli\second} & \SI{7.18}{\second} & \SI{5.53}{\milli\second}  & \SI{119.8}{\milli\second}   \\ 
                GPU & \SI{2.5}{\milli\second}  & \SI{22.9}{\milli\second}& \SI{4.00}{\milli\second}  & \SI{12.53}{\second} \\
                \hline
                \hline
                    & JAX & $\nabla$ JAX && \\
                \hline
                CPU & \SI{48}{\milli\second}   & \SI{364}{\milli\second}&   & \\
                GPU & \SI{887}{\micro\second}   & \SI{10.2}{\milli\second}&   &
                \end{tabular}
                \caption{Runtimes of a single Total Variation regularizer calculation for different Julia versions and JAX. 
                    We used a 3 dimensional Float32 array with size $300 \times 300 \times 300$.  }
                \label{tab:reg}
            \end{table}
            We also compare the performance to an implementation written in Python with the Framework JAX \cite{jax2018github}. 
            The GPU version seems to outperform the Julia version by roughly a factor of $2 - 3$ in both forward and gradient pass.
            However, on the CPU \verb|Tullio.jl| is considerably faster. 
            In general one could always squeeze more performance out of the gradient calculations if the kernels would be hand optimized.
            In our package we especially try avoid that because of the multi-dimensionality it is much more convenient
            to rely on AD and as shown above the performance is usually acceptable.

    \subsection{Full Deconvolution}
        In \autoref{tab:deconv} we show the deconvolution time for a 3D dataset with a size of $512 \times 256 \times 128$\footnote{Dataset was taken from \url{http://bigwww.epfl.ch/deconvolution/data/microtubules/}\cite{deconvlab2}.} 
        The basic settings of the different algorithms and packages are noted in \autoref{tab:deconvprop}.
        The aim of this comparison is not to compare the quality of the results but the computing time to achieve similar reconstructions. 
        The algorithms used in \verb|DeconvOptim.jl| are not fundamentally different to a plain Richardson-Lucy deconvolution and therefore the results will 
        not exceed those results significantly in terms of quality.\\
        One should keep in mind that such performance comparisons are usually biased since we are not
        familiar with all options which should be chosen to achieve best performance.
        The comparison should give a rough overview which runtime is expected for each toolbox.\\
        The different packages used were \verb|ThreeDeconv.jl| \cite{ikoma2018convex}, \verb|Huygens| Professional version 21.04 (Scientific Volume Imaging, The Netherlands, \url{http://svi.nl}), \verb|DeconvolutionLab2| \cite{deconvlab2} and \verb|GenericDeconv|.

        \begin{table}[h]
            \begin{tabular}{l l l l }
                &  CPU in \si{\second}& GPU in \si{\second} & NCC\\ 
            \hline
            \verb|DeconvOptim.jl| & 32.9 & 2.20 & 0.86\\
            \verb|DeconvOptim.jl| - TV & 39.9 & 4.46 & 0.79\\
            Richardson-Lucy (Julia) & 40.0 & 3.05 & 0.74 \\
            \verb|ThreeDeconv.jl|& 1550 & 27.1 & 0.85\\
            \verb|GenericDeconv|  & 140 & 6.32 & 0.86\\
            \verb|GenericDeconv| - TV & 233 & 9.35 & 0.73\\
            \verb|Huygens| & 30.1 & 6.8 & 0.86\\
            \verb|DeconvolutionLab2| - LR& 1640 & & 0.76\\
            \end{tabular}
            \caption{Runtimes of different deconvolution implementations.}
            \label{tab:deconv}
        \end{table}

        For the results of \verb|GenericDeconv| (written in Matlab by Rainer Heintzmann) we used a Poisson loss function with a L-BFGS minimizer 
        and a quadratic non-negativity constraint.
        When the TV regularizer was added the weight was $\lambda = 0.001$.
        The general idea of \verb|GenericDeconv| is very similar to \verb|DeconvOptim.jl|.
        \verb|DeconvolutionLab2| offers a wide variety of different algorithms which however only run on a CPU. 
        In the table we show the results of an 
        iterative Richardson-Lucy deconvolution. The regularizer weight was $\lambda=0.001$.\\
        In \verb|DeconvOptim.jl| we measured the runtime of the default deconvolution routine with a Poisson loss function
        and the runtime of straightforward Richardson-Lucy deconvolution without a regularizer.\\
        \verb|ThreeDeconv.jl| is another Julia deconvolution toolbox for deconvolution of three dimensional data, the regularizer weight was $\lambda=10^{-7}$. 
        Since it is hard to judge the quality of those results with a numerical loss function value, 
        we compare them by the NCC, see \autoref{img:results}.
        The CPU versions were multi threaded (if possible). 
        \verb|DeconvolutionLab2|, \verb|Huygens| and \verb|GenericDeconv| were tested under Windows 10 and not Ubuntu.

        \begin{table}[h]
            \begin{tabular}{l l l}
                & Loss & iterations \\ 
            \hline
            \verb|DeconvOptim.jl| & Poisson & 45 \\
            \verb|DeconvOptim.jl| TV & Poisson & 45\\
            Richardson-Lucy & Poisson & 300 \\
            \verb|ThreeDeconv.jl|  & Gauss+Poisson& 300\\
            \verb|GenericDeconv|  & Poisson & 45\\
            \verb|GenericDeconv| TV & Poisson & 45 \\
            \verb|Huygens| & & \\
            \verb|DeconvolutionLab2| LR & Poisson & 300\\
            \verb|DeconvolutionLab2| LR TV& Poisson & 300 
            \end{tabular}
            \caption{Overview which loss function (noise model) and number of iterations are used.}
            \label{tab:deconvprop}
        \end{table}
        As summary of those performance tests we can state that our Julia implementation is very competitive and especially on the GPU
        it was the fastest of the tested packages. 
        We also see that our routine was faster than a plain Richardson-Lucy mainly due to the much lower number
        of iterations.
        On the CPU \verb|DeconvOptim.jl| is only beaten by the commercial \verb|Huygens| software. 
        Despite \verb|ThreeDeconv.jl| written in Julia it is notably slower in comparison to our implementation.
        \verb|DeconvolutionLab2| is written in Java and uses hand optimized gradients but still its speed is poor in
        comparison to all other packages.
